---
title: 'Splicing influence estimation by Long Reads: SaiLoR'
output:
  html_document: default
  pdf_document: default
---

```{r setup, warning=FALSE, message=FALSE}
library(SaiLoR)
```

```{r, include=FALSE}
library(dplyr)
library(ggplot2)
```


# Running SAILOR 

### Input files 
  * Genome Alignment bam files [minimap2](https://github.com/lh3/minimap2) using parameters `minimap2 -ax splice -u f annotation/genome.fa long_read.fastq.gz | samtools sort -@ 4 -o output.bam - samtools index output.bam`
  * Reference annotation in gtf format. Example file [here](https://github.com/hilgers-lab/SaiLoR/blob/master/inst/exdata/dm6.annot.gtf.gz) 
  * Short read sequencing SJ.out files [STAR](https://github.com/alexdobin/STAR). Example file in [here](https://github.com/hilgers-lab/SaiLoR/blob/master/inst/exdata/short_read_junctions.SJ.out.tab). We recommend to pull SJ.out into a single SJ.out from many experiments and filter by min counts.

# Building junction reference. 

SaiLoR requires a reference junction set to identify couplings. It only analyzes junctions that occur outside the influence of 5' selection (alternative transcription start sites) or 3' selection (alternative polyadenylation). Sailor can build the reference using both a reference annotation and short read sequence data junctions from STAR.

```{r, warning=FALSE, message=FALSE}
# load reference annotation 
annot_path <- system.file("exdata/dm6.annot.gtf.gz", package="SaiLoR")
ref_annot <- rtracklayer::import.gff(annot_path)
# load junctions from STAR 
junction_path <- system.file("exdata/short_read_junctions.SJ.out.tab", package = 'SaiLoR')
# create txf file of reference junctions 
referenece_junctions <- create_reference_junctions( junction_path, min.jcounts = 2 , ref_annot, type="short")
referenece_junctions
```


# Read to junctions assignments 

## Full-length read filtering. 

SaiLoR generates a database with read-to-feature assignments. This process involves three steps occurring within `read_to_junctions()`. The steps include: 1) Identifying only 5'-3' full length read isoforms. This filtering is done with `get_sailor_full_lengths()`, considering only reads spanning an annotated TSS and 3' end. 2) Correcting the junctions found in long reads using reference annotation and short read data in `read_refjunction_correction()`. 3) Last, `make_junction_database()` takes each read and assigns the features it contains: TSS, TES, and the junctions it contains.  

```{r,warning=FALSE, message=FALSE}
bamPath <- system.file("exdata/testBam.bam", package = 'SaiLoR')
exonlinks.counts<- read_to_junctions(bamPath, referenece_junctions, annot_path)
```

# Calculate exon-junction couplings to 5'/3'

SaiLoR calculates exon couplings by generating contingency matrices from exon junctions to promoters with `compute_exon_tss_couplings()` and exon junctions to 3' ends with `compute_exon_3end_couplings()`. It uses multinomial testing with the chi-squared statistic to test the significance of the couplings. To account for possible biases due to low counts, the p-value is simulated using Monte Carlo methods.

```{r,warning=FALSE, message=FALSE}
couplings <- calculate_exon_couplings(exonlinks.counts,  referenece_junctions )
```


# Exploring results 

## Transcription start site couplings with exon usage

Looking closely at Stai gene [FBgn0266521] (https://flybase.org/reports/FBgn0266521). We can observed the couplings of the different junctions and their residuals values. Residual values represent the difference between observed and expected junction of a giving pair in the gene. 

Looking into individual promoters we can notice how the first TSS of Stai `FBgn0266521:P01` it's strongly associated with junctions `6122411-6122697` and `6122854-6122977`, while junctions `6121944-6122004` and `6122191-6122263` are depleted from the same promoter but highly enriched in the 2nd TSS `FBgn0266521:P02`. 

```{r, fig.align='center', fig.height=5, fig.width=6.5}
couplings$TSS.couplingsPerJunction %>%
  filter(grepl("FBgn0266521", pairs_id)) %>% mutate(
    promoter_id = stringr::str_split_fixed(.data$pairs_id, ":" , n = 3)[, 3],
    junction = stringr::str_split_fixed(.data$pairs_id, ":" , n = 3)[, 2]
  ) %>% 
  ggplot(., aes(x = junction, 
                y = residuals, 
                color = promoter_id)) + 
  geom_point(size =3) + 
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 45,vjust = 1,hjust = 1)) + 
  facet_grid(cols = vars(promoter_id)) + 
  scale_color_manual(values =c("#FF355E", "#004F98", "#679267")) + 
  geom_hline(yintercept = 0)
```


## 3'end couplings with exon usage

In contrast tacc gene [FBgn0026620] (https://flybase.org/reports/FBgn0026620). We can observed the couplings related to the usage of specific exon-junction to the 3'end selection of the gene. 

```{r, fig.align='center', fig.height=5, fig.width=6.5}
couplings$TES.couplingsPerJunction %>% 
  filter(grepl("FBgn0026620", pairs_id)) %>% 
  mutate(tes_id=stringr::str_split_fixed(.data$pairs_id,":" ,n = 3)[,3], 
         junction=stringr::str_split_fixed(.data$pairs_id,":" ,n = 3)[,2]) %>% 
  ggplot(., 
         aes(x=junction, 
             y=residuals, 
             color=tes_id)) + 
  geom_point(size=3) + 
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) + 
  facet_grid(cols=vars(tes_id)) + 
  scale_color_manual(values=c("#FF355E","#004F98", "#679267", "blue", "lavender")) + 
  geom_hline(yintercept=0)
```

## Session information

```{r SessionInfo, eval=TRUE, echo=FALSE}
sessionInfo()
```




